---
title: Save/load spatial omics data to/from file
author:
- name: Aaron Lun
  email: luna@gene.com
  affiliation: Data Science and Statistical Computing 
package: alabaster.spatial 
date: "Revised: September 21, 2022"
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Spatial data in DataSetDB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
library(BiocStyle)
self <- Githubpkg("ArtifactDB/alabaster.spatial")
```

# Overview

The `SpatialExperiment` class (from the `r Biocpkg("SpatialExperiment")` package) provides a representation of spatial transcriptomics data that is compatible with Bioconductor's `SummarizedExperiment` ecosystem.
The `r self` package contains methods to save and load `SpatialExperiment` objects into and out of file. 
Check out the `r Githubpkg("ArtifactDB/alabaster.base")` for more details on the motivation and concepts of the **alabaster** framework.

# Quick start

To demonstrate, we'll use the example dataset provided in the `r Biocpkg("SpatialExperiment")` package:

```{r}
library(SpatialExperiment)
example(read10xVisium, echo=FALSE)

# Making the column names unique.
colnames(spe) <- make.unique(colnames(spe))

spe
```

We call the usual `stageObject()` function to save its contents to file:

```{r}
library(alabaster.spatial)

tmp <- tempfile()
dir.create(tmp, showWarnings=FALSE)
meta <- stageObject(spe, dir=tmp, "spe", stage.only=TRUE)
.writeMetadata(meta, tmp)

list.files(tmp, recursive=TRUE)
```

This goes through the usual saving process for `SingleCellExperiment`s, with an additional saving step for the image data (see the `spatial/` subdirectory).
We can then load it back in using the `loadObject()` function:

```{r}
remeta <- acquireMetadata(tmp, "spe/experiment.json")
roundtrip <- loadObject(remeta, tmp)
spe2
plot(imgRaster(getImg(spe2, "section1")))
```

More details on the metadata and on-disk layout are provided in the [schema](https://artifactdb.github.io/BiocObjectSchemas/html/spatial_experiment/v1.html).

# Session info {-}

```{r}
sessionInfo()
```
